Including EW and mixed QCD-EW corrections into PDF fits

### 2020-04-06 paper/sec-introduction.tex
Fix small mistakes in the introduction
- ran a spell checker over the introduction
- use siunitx for units
- Drell-Yan -> Drell--Yan (en dash instead of a hyphen)
- changes dictated by the JHEP author's manual:
  - 'section' instead of 'Sec.' or 'Sect.'
  - 'ref.' instead of 'Ref.'
  - no italicised latin abbreviations

With the recent completion of Run II, the Large Hadron Collider (LHC) has 
accumulated data from an integrated luminosity of approximately 
150~fb$^{-1}$~\cite{Mangano:2020icy}. This represents only a small fraction of 
the anticipated 3000 fb$^{-1}$ that will eventually be recorded in the 
forthcoming twenty years of LHC operation. Nevertheless the statistical 
uncertainty of the data has already shrunk to unprecedentedly small values,
typically 1\% or less, a fact that will allow for precision tests of the 
Standard Model (SM) and for indirect searches of New Physics only if 
theoretical predictions become comparatively precise. This entails the 
computation of additional higher-order contributions to the fixed-order 
coupling-constant perturbative expansion, on the one hand, and an increasingly 
sophisticated determination of the Parton Distribution Functions (PDFs) of the 
proton~\cite{Gao:2017yyd}, on the other hand.

In the first respect, because Quantum Chromodynamics (QCD) dominates the 
interactions occurring within colliding protons at the LHC, much effort 
has been devoted to the computation of higher-order QCD corrections: 
fully-differential next-to-leading order (NLO) results, possibly matched to the
parton shower, are currently automated in various general-purpose
Monte Carlo generators~\cite{Gleisberg:2008ta,Alwall:2014hca,Bellm:2015jjp}
(see also~\cite{Buckley:2011ms} for a review),
while an increasing number of next-to-next-to-leading order (NNLO) predictions 
are becoming available for processes with various degrees of inclusiveness
(see {\it e.g.}~\cite{Amoroso:2020lgh} and references therein). The computation
of higher-order corrections in the Electroweak (EW) and mixed QCD+EW theory has 
witnessed a comparatively less comprehensive progress. Nevertheless frameworks 
were developed in which the QCD and EW couplings are simultaneously treated as 
small parameters in the perturbative expansion, and the computation of 
theoretical predictions, accurate to NLO in both (including mixed-coupling 
QCD+EW terms), is automated~\cite{Biedermann:2017yoi,Frederix:2018nkq}.

In the second respect, contemporary PDF
sets~\cite{Harland-Lang:2014zoa,Ball:2017nwa,Hou:2019efy}
incorporate a significant amount of LHC data, which is analysed with NNLO QCD 
theory by default. No EW corrections are systematically included in the 
theoretical description of the experimental observables to which PDFs are 
optimised, except for QED effects if a photon PDF is 
determined~\cite{Schmidt:2015zda,Bertone:2017bme,Harland-Lang:2019pla}. 
The resulting relative PDF uncertainy --- which accounts only for the 
uncertainty of the data and of residual methodology inefficiencies inherent to 
each PDF determination --- can be as low as 1\% at the EW 
scale~\cite{Ball:2017nwa}. Theoretical uncertainties, possibly of comparable 
size ({\it e.g.} from missing higher-order terms in the QCD perturbative 
expansion), have started to be included in a PDF determination and represented 
into PDF uncertainties only very 
recently~\cite{AbdulKhalek:2019bux,AbdulKhalek:2019ihb}. This has led to a 
noticeable improvement of PDF accuracy, in terms of an overall
better statistical quality of the fit, at the price of a modest deterioration 
of PDF precision, in terms of a slight inflation of PDF uncertainties.

@@

The systematic inclusion of EW corrections in a PDF determination entails
the solution of two separate problems. First, a problem of efficiency:
fast interpolation grids should be constructed, whereby partonic matrix 
elements, accurate to NLO QCD+EW, are precomputed in such a way that the 
numerical convolution with generic input PDFs can be efficiently approximated
by means of interpolation techniques. Such grids are a fundamental ingredient
in any PDF fit, where the evaluation of the hadronic cross section needs to 
be performed a large number of times. While two default formats exist for
these grids, {\sc APPLgrid}~\cite{Carli:2010rw} and 
{\sc fastNLO}~\cite{Kluge:2006xs}, none of them supports the inclusion of EW
corrections nor the interface to a Monte Carlo generator accurate to 
NLO QED+EW. Second, a problem of consistency: the way in which EW effects may
(or may not) be folded into the data varies across different experimental 
analyses, a fact that challenges their consistent theoretical interpretation. 
Examples are the subtraction of background processes which should not be 
considered as such ({\it e.g.} the $t$-channel photon-induced component in 
neutral-current Drell-Yan, which is not a separate processes beyond the leading
order) or of just a part of the EW effects ({\it e.g.} multiple-photon 
radiation from light particles in the final state of neutral- or charged-current
Drell-Yan, especially with electrons). Be that as it may, if EW effects are 
systematically included in theoretical predictions, they should not be included 
in the data, otherwise they will be double counted.

In this paper we address the first of these two problems. Specifically, 
we develop {\sc PineAPPL}, a new library that provides a direct interface 
from parton-level Monte Carlo event generators accurate up to NLO QCD+EW to
fast interpolation grids. The library supports variations of the factorisation
and renormalisations scales, resummation, and matching with a photon- and/or
parton-shower. The library is included by default in 
{\sc MadGraph5\_aMC@NLO} ({\sc mg5\_aMC@NLO} henceforth) within which it has 
been designed and tested. However, it is suitable for use as a plugin in any 
Monte Carlo generator that supports NLO QCD+EW computations, such as 
{\sc SHERPA}~\cite{Biedermann:2017yoi}. In this respect, {\sc PineAPPL}
supersedes the {\sc aMCfast} interface~\cite{Bertone:2014zva} and complements
the {\sc MCgrid} interface~\cite{DelDebbio:2013kxa,Bothmann:2015dba}.

The paper is organised as follows. In Sect.~\ref{sec:pineappl} we introduce
{\sc PineAPPL}, we describe its features, we illustrate its operation, and we
assess its performance. In Sect.~\ref{sec:results}, by means of 
{\sc mg5\_AMC@NLO} and {\sc PineAPPL}, we compute fast interpolation grids for 
a representative set of data that is typically included in a fit of PDFs.
In Sect.~\ref{sec:doublecounting} we use these results to formulate more 
comprehensively the second problem sketched above, the solution of which, 
however, remains beyond the scope of the current work. We provide our
conclusions and an outlook in Sect.~\ref{sec:conclusion}.

>>>>

With the recent completion of Run II, the Large Hadron Collider (LHC) has 
accumulated data from an integrated luminosity of approximately 
\SI{150}{\per\femto\barn}~\cite{Mangano:2020icy}. This represents only a small fraction of
the anticipated \SI{3000}{\per\femto\barn} that will eventually be recorded in the
forthcoming twenty years of LHC operation. Nevertheless the statistical 
uncertainty of the data has already shrunk to unprecedented small values,
typically \SI{1}{\percent} or less, a fact that will allow for precision tests of the
Standard Model (SM) and for indirect searches of New Physics only if 
theoretical predictions become comparatively precise. This entails the 
computation of additional higher-order contributions to the fixed-order 
coupling-constant perturbative expansion, on the one hand, and an increasingly 
sophisticated determination of the Parton Distribution Functions (PDFs) of the 
proton~\cite{Gao:2017yyd}, on the other hand.

In the first respect, because Quantum Chromodynamics (QCD) dominates the 
interactions occurring within colliding protons at the LHC, much effort 
has been devoted to the computation of higher-order QCD corrections: 
fully-differential next-to-leading order (NLO) results, possibly matched to the
parton shower, are currently automated in various general-purpose
Monte Carlo generators~\cite{Gleisberg:2008ta,Alwall:2014hca,Bellm:2015jjp}
(see also ref.~\cite{Buckley:2011ms} for a review),
while an increasing number of next-to-next-to-leading order (NNLO) predictions 
are becoming available for processes with various degrees of inclusiveness
(see e.g.\ ref.~\cite{Amoroso:2020lgh} and references therein). The computation
of higher-order corrections in the electroweak (EW) and mixed QCD+EW theory has
witnessed a comparatively less comprehensive progress. Nevertheless frameworks 
were developed in which the QCD and EW couplings are simultaneously treated as 
small parameters in the perturbative expansion, and the computation of 
theoretical predictions, accurate to NLO in both (including mixed-coupling 
QCD+EW terms), is automated~\cite{Biedermann:2017yoi,Frederix:2018nkq}.

In the second respect, contemporary PDF
sets~\cite{Harland-Lang:2014zoa,Ball:2017nwa,Hou:2019efy}
incorporate a significant amount of LHC data, which is analysed with NNLO QCD 
theory by default. No EW corrections are systematically included in the 
theoretical description of the experimental observables to which PDFs are 
optimised, except for QED effects if a photon PDF is 
determined~\cite{Schmidt:2015zda,Bertone:2017bme,Harland-Lang:2019pla}. 
The resulting relative PDF uncertainty --- which accounts only for the
uncertainty of the data and of residual methodology inefficiencies inherent to 
each PDF determination --- can be as low as 1\% at the EW 
scale~\cite{Ball:2017nwa}. Theoretical uncertainties, possibly of comparable 
size (e.g.\ from missing higher-order terms in the QCD perturbative
expansion), have started to be included in a PDF determination and represented 
into PDF uncertainties only very 
recently~\cite{AbdulKhalek:2019bux,AbdulKhalek:2019ihb}. This has led to a 
noticeable improvement of PDF accuracy, in terms of an overall
better statistical quality of the fit, at the price of a modest deterioration 
of PDF precision, in terms of a slight inflation of PDF uncertainties.

@@

The systematic inclusion of EW corrections in a PDF determination entails
the solution of two separate problems. First, a problem of efficiency:
fast interpolation grids should be constructed, whereby partonic matrix 
elements, accurate to NLO QCD+EW, are precomputed in such a way that the 
numerical convolution with generic input PDFs can be efficiently approximated
by means of interpolation techniques. Such grids are a fundamental ingredient
in any PDF fit, where the evaluation of the hadronic cross section needs to 
be performed a large number of times. While two default formats exist for
these grids, {\sc APPLgrid}~\cite{Carli:2010rw} and 
{\sc fastNLO}~\cite{Kluge:2006xs}, none of them supports the inclusion of EW
corrections nor the interface to a Monte Carlo generator accurate to 
NLO QED+EW. Second, a problem of consistency: the way in which EW effects may
(or may not) be folded into the data varies across different experimental 
analyses, a fact that challenges their consistent theoretical interpretation. 
Examples are the subtraction of background processes which should not be 
considered as such (e.g.\ the $t$-channel photon-induced component in
neutral-current Drell--Yan, which is not a separate processes beyond the leading
order) or of just a part of the EW effects (e.g.\ multiple-photon
radiation from light particles in the final state of neutral- or charged-current
Drell--Yan, especially with electrons). Be that as it may, if EW effects are
systematically included in theoretical predictions, they should not be included 
in the data, otherwise they will be double counted.

In this paper we address the first of these two problems. Specifically, 
we develop {\sc PineAPPL}, a new library that provides a direct interface 
from parton-level Monte Carlo event generators accurate up to NLO QCD+EW to
fast interpolation grids. The library supports variations of the factorisation
and renormalisation scales, resummation, and matching with a photon- and/or
parton-shower. The library is included by default in 
{\sc MadGraph5\_aMC@NLO} ({\sc mg5\_aMC@NLO} henceforth) within which it has 
been designed and tested. However, it is suitable for use as a plugin in any 
Monte Carlo generator that supports NLO QCD+EW computations, such as 
{\sc SHERPA}~\cite{Biedermann:2017yoi}. In this respect, {\sc PineAPPL}
supersedes the {\sc aMCfast} interface~\cite{Bertone:2014zva} and complements
the {\sc MCgrid} interface~\cite{DelDebbio:2013kxa,Bothmann:2015dba}.

The paper is organised as follows. In section~\ref{sec:pineappl} we introduce
{\sc PineAPPL}, we describe its features, we illustrate its operation, and we
assess its performance. In section~\ref{sec:results}, by means of
{\sc mg5\_AMC@NLO} and {\sc PineAPPL}, we compute fast interpolation grids for 
a representative set of data that is typically included in a fit of PDFs.
In section~\ref{sec:doublecounting} we use these results to formulate more
comprehensively the second problem sketched above, the solution of which, 
however, remains beyond the scope of the current work. We provide our
conclusions and an outlook in section~\ref{sec:conclusion}.
